---
title: "How to Make a Blog Post: a Brief Introduction"
description: |
 Learn how to contribute to the wiki section of the Ecological Dynamics Department webpage. If you think you have an important information, script, package or just a pieace of code that might be interesting for your colleagues you can create easily a blog post and add this to the wiki section of the webpage.
categories:
  - tutorial
  - workflow
author:
  - name: Moritz Wenzler-Meya
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/EcoDynIZW
date: 2023-06-14
output:
  distill::distill_article:
    self_contained: false
    toc: true
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You have a nice peace of code? You have made a cool package? You have something to share within (or even outside) our department?
  
You can share it with a blogpost on our ecodynizw Website 
  
# Step 1 
  - Clone the [GitHub repository](https://github.com/EcoDynIZW/EcoDynIZW.github.io)
  - Find the _posts_template.Rmd and start with filling all the information/code/images into the template
  - please describe every task briefly 
  
# Step 2
  - knit the post and check if the knitted document looks as desired
  - push and commit everything to github 
  - create a pull request and select a reviewer

# Step 3 
  - wait for the comments of the reviewer and if there are any include them and push again the script
  - the reviewer will merge the pull request and shortly your post will appear on the website under coding posts


```{r, eval=FALSE, echo=T}
post_fun <- function(post_name){
  if(dir.exists(here::here("_posts", post_name)) == FALSE){
  dir.create(here::here("_posts", post_name))
  }
  file.copy("_template_posts.Rmd", here::here("_posts", post_name, paste0(post_name, ".rmd")))
}
```

# general tips for posting

- please use the spell check before knitting and pushing the post

- please check if the code is running (in a reasonable time) as we have to rebuild the page sometimes

- examples for styling like `plot` for functions examples

- maybe show how to hide code? Or how to add pseudo code? 

- prepare/update the "_post_template.rmd"


