---
title: "How to Make a Blog Post: a Brief Introduction"
description: |
 Learn how to contribute to the wiki section of the Ecological Dynamics Department webpage. If you think you have an important information, script, package or just a pieace of code that might be interesting for your colleagues you can add a blog post to the wiki section of our webpage.
categories:
  - tutorial
  - workflow
author:
  - name: Moritz Wenzler-Meya
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/EcoDynIZW
date: 2023-06-14
output:
  distill::distill_article:
    self_contained: false
    toc: true
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You have a nice piece of code? You have developed a cool package? You have something to share within (or even outside) our department?
  
You can share it with a blogpost on our ecodynizw Website by folllowing these simple steps:
  
### Step 1 

  - Clone the [GitHub repository](https://github.com/EcoDynIZW/EcoDynIZW.github.io)
  - Find the _posts_template.Rmd and start with filling all the information/code/images into the template
  - please describe every task briefly 
  
### Step 2

  - knit the post and check if the knitted document looks as desired
  - push and commit everything to github 
  - create a pull request and select a reviewer

### Step 3 

  - wait for the comments of the reviewer and if there are any include them and push again the script
  - the reviewer will merge the pull request and shortly your post will appear on the website under coding posts

```{r, eval=FALSE, echo=TRUE}
post_fun <- function(post_name){
  if(dir.exists(here::here("_posts", post_name)) == FALSE){
  dir.create(here::here("_posts", post_name))
  }
  file.copy("_template_posts.Rmd", here::here("_posts", post_name, paste0(post_name, ".rmd")))
}
```

### General tips for posting

- Please use the spell check before knitting and pushing the post.

- Please check if the code is running (in a reasonable time) as we have to rebuild the page sometimes.

- Examples for styling like `plot` for functions examples.

- TODO: maybe show how to hide code? Or how to add pseudo code?

- TODO: prepare/update the "_post_template.rmd"


