---
title: "How to make a blogpost - a brief introduction"
description: |
 Learn how to contribute to the wiki section of the Ecological Department webpage. If you think you have an important information, script, package or just a pieace of code that might be intersesting for your collegues you can create easily a blog post and add this to the wiki section of the webpage.
categories:
  - tutorial
  - workflow
author:
  - name: Moritz Wenzler-Meya
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/EcoDynIZW
date: 2023-02-20
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- why contributing?
  You have a nice peace of code? You made a cool package?

- how to contribute?
  You have to clone the [EcoDynIZW.github.io](x-github-client://openRepo/https://github.com/EcoDynIZW/EcoDynIZW.github.io) repository, create a post by using the _post_template.rmd and make a merge request when done.
  You have to create a folder with the name of the post and save the rmd in there (maybe we create a function for that, like our own post wiki function or so).
  
```{r}
distill::create_post()

post_fun <- function(post_name){
  if(dir.exists(here::here("_posts", post_name)) == FALSE){
  dir.create(here::here("_posts", post_name))
  }
  file.copy("_template_posts.Rmd", here::here("_posts", post_name, paste0(post_name, ".rmd")))
}

```

  

- Please check before for spelling mistakes and if the knitting is working easily, because sometimes we have to rerun the knitting.

- examples for styling like `plot` for functions examples

- maybe show how to hide code? Or hot to make pseudo code? 

- prepare/update the "_post_template.rmd"


