---
title: "Simplify Workflows of D6 Research Projects"
description: |
  Learn how to use the {d6} package to follow the project workflow within the department “Ecological Dynamics” at the Leibniz Institute for Zoo and Wildlife Research. The package functionality allows you to set up a standardized folder structure, to use templates for standardized reports and provides some helpful utility functions.
preview: https://raw.githubusercontent.com/EcoDynIZW/EcoDynIZW.github.io/main/img/wiki/hex-d6.png 
categories:
  - tutorial
  - rstats
  - workflow
  - data management
author:
  - name: Cedric Scherer
    url: https://cedricscherer.com
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/
date: 2020-12-09
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


The [`{d6}` package](https://github.com/EcoDynIZW/d6) aims to simplify workflows of our D6 research projects by providing a standardized folder structure incl. version control, Rmarkdown templates, and other utilities.

There are four main functionalities:

1.  [Create standardized project directories with
    `new_project()`](#create-project-directory)
2.  [Install a set of common packages with
    `install_d6_packages()`](#install-common-packages)
3.  [Provide custom Rmarkdown templates via
    `File > New File > Rmarkdown... > From Template`](#use-custom-rmarkdown-templates)
4.  [Render all Rmarkdown documents to ./docs/report with
    `render_all_reports()` or
    `render_report()`](#render-rmarkdown-files-to-reports)

## Installation

The package is not on CRAN and needs to be installed from GitHub. To do
so, open Rstudio and run the following two lines in the console. In case
the `{devtools}` package is already installed, skip that step.

```{r}
install.packages("devtools")
devtools::install_github("EcoDynIZW/d6")
```

(Note: If you are asked if you want to update other packages either press "No" (option 3) and continue or update the packages before running the install command again.)



## Create Project Directory {#create-project-directory}

Run the function `new_project()` to create a new project. This will
create a standardized directory with all the scaffolding we use for all
projects in our department. It also add several files needed for
documentation of your project.

To start a new project in the current working directory, simply run:

```{r}
d6::new_project("unicornus_wl_sdm_smith_j")
```

*Please give your project a unique and descriptive name:
**species\_country\_topic\_name***

For example, when John Smith is developing a species distribution models
for unicorns in Wonderland, a descriptive title could be:
`unicornus_wl_sdm_smith_j`. Please use underscores and the
[international Alpha-2 encoding for
countries](https://www.iban.com/country-codes).

The full scaffolding structure created in the root folder (here
`unicornus_wl_sdm_smith_j`) is the following:

```{verbatim}
. 
└── unicornus_wl_sdm_smith_j
    ├── .Rproj.user         —  Rproject files
    ├── data-raw            —  raw data (tabular data in root folder)
    │    └── geo-raw        —  raw spatial data
    ├── docs                —  documents
    │   ├── admin           —  adminstrative docs, e.g. permits 
    │   ├── literature      —  literature used for parameterization + ms
    │   ├── manuscript      —  manuscript drafts (main + supplement)
    │   ├── presentations   —  talks and poster presentations
    │   └── reports         —  rendered reports
    ├── output              —  everything that is computed (except plots)
    │   ├── data-proc       —  processed tabular data
    │   └── geo-proc        —  processed spatial data
    ├── plots               —  plot output
    ├── R                   —  scripts
    │   ├── 00_start.R      —  first script to run
    │   └── XX_submit.R     —  final script to run
    ├── .gitignore          —  contains which files to ignore for version control
    ├── .Rbuildignore       —  contains which files to ignore for package builds
    ├── DESCRIPTION         —  contains project details and package dependencies
    ├── NAMESPACE           —  contains context for R objects
    └── project.Rproj       —  Rproject file: use to start your project
```

### Use A Custom Root Directory

You do not need to change the working directory first---you can also
specify a path to a custom root folder in which the new project folder
is created:

```{r}
## both work:
d6::new_project("unicornus_wl_sdm_smith_j", path = "absolute/path/to/the/root/folder")
## or:
d6::new_project("unicornus_wl_sdm_smith_j", path = "absolute/path/to/the/root/folder/")
```

The resulting final directory of your project would be
`absolute/path/to/the/root/folder/unicornus_wl_sdm_smith_j`.

### Use Version Control

If you want to create a GitHub repository for the project at the same
time, use instead:

```{r}
d6::new_project("unicornus_wl_sdm_smith_j", github = TRUE)
```

By default, the visibility of the GitHub repository is set to “private”
but you can also change that:

```{r}
d6::new_project("unicornus_wl_sdm_smith_j", github = TRUE, private_repo = FALSE)
```

Note that to create a GitHub repo you will need to have configured your
system as explained [here](https://gist.github.com/Z3tt/3dab3535007acf108391649766409421).


### Setup without Geo Directories

If your project does not (or will not) contain any spatial data, you can
prevent the creation of the directories `geo-raw` and `geo-proc` by
setting `geo` to `FALSE`:

```{r}
d6::new_project("unicornus_wl_sdm_smith_j", geo = FALSE)
```

### Add Documentation to Your Project

After you have set up your project directory, open the file `00_start.R`
in the `R` folder. Add the details of your project, fill in the readme,
add a MIT license (if needed) and add package dependencies.



## Install Common Packages {#install-common-packages}

You can install the packages that are most commonly used in our
department via `install_d6_packages()`:

```{r}
d6::install_d6_packages()
```

Note that this function is going to check preinstalled versions and will
only install packages that are not installed with your current R
version.

Again, there is an arguement `geo` so you can decide if you want to
install common geodata packages as well (which is the default). If you
are not intending to process geodata, set `geo` to `FALSE`:

```{r}
d6::install_d6_packages(geo = FALSE)
```

The default packages that are going to be installed are:

```{verbatim}
tidyverse (tibble, dplyr, tidyr, ggplot2, readr, forecats, stringr, purrr), lubridate, here, vroom, patchwork, usethis
```

The following packages will be installed in case you specify
`geo = TRUE`:

```{verbatim}
rgdal, geos, raster, sp, sf, tmap
```


## Use Custom Rmarkdown Templates {#use-custom-rmarkdown-templates}

The package also provides several templates for your scripts. In
Rstudio, navigate to `File > New File > RMarkdown... > Templates` and
choose the template you want to use. All templates come with a
preformatted YAML header and chunks for the setup.

The following templates are available for now:

-   *EcoDynIZW Basic*: Template for a basic Rmarkdown research report
    including bits of codes and comments to get started
-   *EcoDynIZW Minimal*: Template for an Rmarkdown research report
    (almost empty)



## Render Rmarkdown Files to Reports {#render-rmarkdown-files-to-reports}

The `render_*()` functions take care of knitting your Rmarkdown files
into HTML reports. The functions assume that your .Rmd files are saved
in the `R` directory or any subdirectory, and will store the resulting
.html files in the according directory, namely `./docs/reports/`.

You can render all .Rmd files that are placed in the `R` directory and
subdirectories in one step:

```{r}
d6::render_all_reports()
```

You can also render single Rmarkdown documents via `render_report()`:

```{r}
d6::render_report("my-report.Rmd")
d6::render_report("notsurewhybutIhaveasubfolder/my-report.Rmd")
```


