---
title: "camtrapR"
description: |
  Manage camera trap data in R
preview: https://cran.r-project.org/web/packages/camtrapR/index.html
author:
  - name: Moritz Wenzler-Meya
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/EcoDynIZW
date: 12-09-2020
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

[camtrapR](https://github.com/jniedballa/camtrapR): Manage camera trap data in R

camtrapR is a package for camera trap data management in R.

## Installation

You can install the release version of camtrapR from CRAN:

```{r, eval=FALSE, echo=TRUE}
install.packages("camtrapR")
```

or the latest GitHub version (containing recent changes and new features) via:

```{r, eval=FALSE, echo=TRUE}
# install.packages("remotes")
# install.packages("R.rsp")
remotes::install_github("jniedballa/camtrapR", build_vignettes = TRUE)
```

For installation from GitHub make sure package R.rsp is available. build_vignettes = TRUE is necessary since v 2.1.1. due to the new static vignettes.

You can also install a previous version (e.g. if a recent update broke the code you need to run urgently) via:

```{r, eval=FALSE, echo=TRUE}
remotes::install_version(package = "camtrapR", version = "2.0.3")
```

## Exiftool

Numerous important camtrapR functions read EXIF metadata from JPG images (and videos). This is done via Exiftool, a free and open-source sofware tool developed by Phil Harvey and available for Windows, MacOS and Linux.

To make full use of camtrapR, you will need Exiftool on your system. You can download it from the Exiftool homepage. Please follow the installation instruction in vignette 1 and the Exiftool website.

You may not need Exiftool if you do not work with image files, but only use camtrapR to create input for occupancy or spatial capture-recapture models from existing record tables.
Help and Support
Google Group

There is a Google Group for all questions and problems regarding camtrapR. Please feel free to join the discussion there.
GitHub issues

Alternatively, you can report bugs and problems as issues on GitHub (here).
Further reading

See the Article in Methods in Ecology and Evolution for an overview of the package. The five vignettes provide examples for the entire workflow.
Citation

## Citation

Jürgen Niedballa, Rahel Sollmann, Alexandre Courtiol, Andreas Wilting (2016). camtrapR: an R package for efficient camera trap data management. Methods in Ecology and Evolution 7(12), 1457-1462, [https://doi.org/10.1111/2041-210X.12600](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12600) 

or see 

```{r, eval=FALSE, echo=TRUE}
citation("camtrapR")
```


