---
title: "d6geodata - a package to manage spatial data in the Ecodyn Department of the IZW"
description: |
 This package provides functions to show and get geodata. 
  To download data from the fisbroker database and several functions for the geodata manager to provide these datasets. 
categories:
  - tutorial
  - spatial
  -
  
author:
  - name: Moritz Wenzler-Meya
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/EcoDynIZW
date: 2023-01-31
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The [`{d6geodata}` package](https://github.com/EcoDynIZW/d6geodata) provides several functions for the members of the Ecological Department of the IZW. Some functions can be used without access to our resources such as *download_fisbroker_wfs* and *download_fisbroker_atom*.

```{r install, eval=3}
devtools::install_github("EcoDynIZW/d6geodata")
library(d6geodata)
```

Functions for all: 

- download_fisbroker_wfs
- download_fisbroker_atom


For using these two functions you have to go to the fisbroker data base and select a layer you want to download. 
```{r example}

```

Functions for internal use only:

- geo_overview
- get_geodata


```{r plot, echo=FALSE}

```

