---
title: "Manage Spatial Data from Our Geodata Archive"
description: |
  Learn how to use the {d6geodata} R package that provides functions for accessing data from the Geodata archive of the Department of Ecological Dynamics. The two functions `geo_overview()` and `get_geodata()` are the main components for all members of our Department. Several other functions are within this package but only meant to be used by the Geodata Manager of the Department.
categories:
  - tutorial
  - spatial
  - rstats
author:
  - name: Moritz Wenzler-Meya
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/EcoDynIZW
date: 2023-03-03
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The [`{d6geodata}` package](https://github.com/EcoDynIZW/d6geodata) aims to access the data from the Geodata archive of the EcoDyn Department for members only!

The two main functions are:

- `geo_overview()` 
- `get_geodata()`


```{r install, eval=3}
remotes::install_github("EcoDynIZW/d6geodata")
library(d6geodata)
```

If you want to work with geodata that is already stored in our Geodata archive you have two options: 

1. Go to the EcoDynIZW Website, click on wikis and select [Geodata](https://ecodynizw.github.io/geodata.html). There you find several spatial data sets with respective metadaat and visualizations. In the metadata section, you'll find the information . To donwload the data, cope the `folder_name` information provided in the metadata  and use it as an input in the  `get_geodata()` function from our `{d6geodata}` to get the data from our PopDynCloud. Another option is the function called `geo_overview()`. There you can select which data and from which location you want to have a list of data.

If you run the function `geo_overview` you have to decide if you want to see the raw or processed data by typing 1 for raw and 2 for processed data. Afterwards, you have to decide if you want to see the main (type 1) folders (the regions or sub-regions we have data from) or the sub (type 2) folders (the actually data we have in each region).

## Example 1: Main Folder

```{r example geo_overview, eval=FALSE}
d6geodata::geo_overview(path_to_cloud = "E:/PopDynCloud")
Raw or processed data: 

1: raw
2: processed

Auswahl: 2
choose folder type: 

1: main
2: sub

Auswahl: 1
[1] "atlas" "BB_MV_B" "berlin" "europe" "germany" "world"
```

## Example 2: Sub Folder

```{r example geo_overview two, eval=FALSE}
d6geodata::geo_overview(path_to_cloud = "E:/PopDynCloud")
Raw or processed data: 

1: raw
2: processed

Auswahl: 2
choose folder type: 

1: main
2: sub

Auswahl: 2
$atlas
[1] "distance-to-human-settlements_atlas_2009_1m_03035_tif"
[2] "distance-to-kettleholes_atlas_2022_1m_03035_tif"      
[3] "distance-to-rivers_atlas_2009_1m_03035_tif"           
[4] "distance-to-streets_atlas_2022_1m_03035_tif"          
[5] "landuse_atlas_2009_1m_03035_tif"                      

$BB_MV_B
[1] "_archive" "_old_not_verified" "dist_path_bb_agroscapelabs"
[4] "scripts"                   

$berlin
 [1] "_old_not_verified"                            
 [2] "corine_berlin_2015_20m_03035_tif"            
 [3] "distance-to-paths_berlin_2022_100m_03035_tif" 
 [4]  "green-capacity_berlin_2020_10m_03035_tif"    
 [5] "imperviousness_berlin_2018_10m_03035_tif"     
 [6]  "light-pollution_berlin_2021_100m_03035_tif"  
 [7] "light-pollution_berlin_2021_10m_03035_tif"    
 [8]  "motorways_berlin_2022_100m_03035_tif"        
 [9] "noise-day-night_berlin_2017_10m_03035_tif"    
[10]  "population-density_berlin_2019_10m_03035_tif"
[11] "template-raster_berlin_2018_10m_03035_tif"    
[12] "tree-cover-density_berlin_2018_10m_03035_tif"

$europe
[1] "imperviousness_europe_2018_10m_03035_tif"

$germany
 [1] "_old_not_verified"                                          
 [2] "distance-to-motorway-rural-road_germany_2022_100m_03035_tif"
 [3] "distance-to-motorways_germany_2022_100m_03035_tif"          
 [4] "distance-to-paths_germany_2022_100m_03035_tif"              
 [5] "distance-to-roads-paths_germany_2022_100m_03035_tif"        
 [6] "distance-to-roads_germany_2022_100m_03035_tif"              
 [7] "distance_to_paths_germany_2022_100m_03035_tif"              
 [8] "motoroways_germany_2022_03035_osm_tif"                      
 [9] "motorway-rural-road_germany_2022_100m_03035_tif"            
[10] "motorways_germany_2022_100m_03035_tif"                      
[11] "paths_germany_2022_100m_03035_tif"                          
[12] "Roads-germany_2022_100m_03035_tif"                          
[13] "roads_germany_2022_100m_03035_tif"                          
[14] "tree-cover-density_germany_2015_100m_03035_tif"             

$world
character(0)
```


Now you can copy the name of one of the layers and paste it into the `get_geodata()` function

```{r example get_geodata}
corine <-
  d6geodata::get_geodata(
    data_name = "corine_berlin_2018_20m_03035_tif",
    path_to_cloud = "E:/PopDynCloud",
    download_data = FALSE
  )
```

If you set `download_data = TRUE` the data will be download and copied to your `data-raw` folder. If the `data-raw` folder doesn't exist, it will be created.

If you want to download more than one file, you can simply use `lapply()` and add multiple file names like this:

```{r example lapply}
data_list <-
  lapply(
    c(
      "corine_berlin_2018_20m_03035_tif",
      "motorways_berlin_2022_100m_03035_tif"
    ),
    FUN = function(x) {
      d6geodata::get_geodata(
        data_name = x,
        path_to_cloud = "E:/PopDynCloud",
        download_data = FALSE
      )})
```


#### Additional functions

The three functions `plot_binary_map()`, `plot_qualitative_map()` and plot `plot_quantitative_map()` can be used to plot raster data with the respective color schemes we used for the Geodata wiki page (note that this function works only for raster data).

```{r plotting functions, eval=FALSE}
plot_binary_map(tif = tif)
plot_qualitative_map(tif = tif)
plot_quantitative_map(tif = tif)
```

#### Example plot

```{r}
library(d6geodata)
plot_qualitative_map(tif = corine)
```



