---
title: "NLMR"
description: |
  NLMR is an R package for simulating neutral landscape models (NLM). Designed to be a generic framework like NLMpy, it leverages the ability to simulate the most common NLM that are described in the ecological literature.
preview: https://raw.githubusercontent.com/EcoDynIZW/EcoDynIZW.github.io/main/img/hex-nlmr.png
author:
  - name: ropensci
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/EcoDynIZW
date: 2022-09-12
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

[NLMR package](https://github.com/ropensci/NLMR) builds on the advantages of the raster package and returns all simulation as RasterLayer objects, thus ensuring a direct compatibility to common GIS tasks and a flexible and simple usage. Furthermore, it simulates NLMs within a self-contained, reproducible framework.

Each neutral landscape models is simulated with a single function (all starting with `nlm_`) in NLMR, e.g.:


```{r, eval=FALSE, echo=TRUE}
random_cluster <- NLMR::nlm_randomcluster(nrow = 100,
                                          ncol = 100,
                                          p    = 0.5,
                                          ai   = c(0.3, 0.6, 0.1),
                                          rescale = FALSE)

random_curdling <- NLMR::nlm_curds(curds = c(0.5, 0.3, 0.6),
                                   recursion_steps = c(32, 6, 2))


midpoint_displacememt <- NLMR::nlm_mpd(ncol = 100,
                                       nrow = 100,
                                       roughness = 0.61)
```

NLMR supplies 15 NLM algorithms, with several options to simulate derivatives of them. The algorithms differ from each other in spatial auto-correlation, from no auto-correlation (random NLM) to a constant gradient (planar gradients).

