---
title: "Raster Operations"
description: |
  Learn how to obtain a template raster from the `{d6geodata}` package, expanding the raster in different ways, rasterizing vector data, and extracting information from a raster using vector data. The methods described involve the use of packages like `{terra}` and `{sf}`, and include examples and plots to illustrate the processes. Additionally, we show the use of the {tidyterra} package for plotting raster data with ggplot.
  
preview: ./img/wiki/preview-rastertemplate.png
categories:
  - tutorial
  - spatial
  - sf
  - terra
author:
  - name: Moritz Wenzler-Meya
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/
date: 2024-03-11
output:
  distill::distill_article:
    self_contained: false
    toc: true
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) ## keep this, add other settings if needed
```

## Setup {#setup}

Install the `{d6geodata}` library and load all necessary libraries. 

```{r}
#remotes::install_github("EcoDynIZW/d6geodata")
library(d6geodata)
library(terra)
library(ggplot2)
library(sf)
library(patchwork)
library(dplyr)

theme_set(
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8))
) 
```


## Raster Template {#template}

First get the template raster. The origin is the extent of Berlin with the projection laea (EPSG:3035).

```{r template}
temp <- d6geodata::temp_ras(1000) # resolution is 1000m
values(temp) <- 1 # set a dummy value for visualization purposes 

temp
```

We load Berlin districts data set from our cloud to show the borders of Berlin.  

```{r berlin borders}
berlin <- d6geodata::get_geodata(data_name = "districs_berlin_2022_poly_03035_gpkg",
                                 path_to_cloud = "E:/PopDynCloud",
                                 download_data = FALSE) %>% 
  st_union() %>% # using `st_union()` to combine  
  st_cast("POLYGON")

p_berlin <- geom_sf(data = berlin, alpha = 0, color = "black", lwd = 0.7)
```


```{r plot template}
p1 <- d6geodata::plot_qualitative_map(temp) # you can use this build-in plot function from the `{d6geodata}` package 
 
p1 + p_berlin
```

## Expand Raster {#expand}

### Expand On All Sides

If you want to expand the raster on all size (similar to a buffer around a polygon) you can just add cells to all sides. You can do this by adding an integer number in the y argument in the function `extend` of the `{terra}` package.

```{r}
temp_exp_all <- terra::extend(x = temp, # template raster as input
                              y = 10, # number of cells to add
                              fill = 2) # value for new cells

p2 <- d6geodata::plot_qualitative_map(temp_exp_all)
p2 + p_berlin
```

### Expand On Two Sides

In this case you can add cells on two opposite sides.

```{r}
temp_exp_tb_sides <- terra::extend(x = temp,
                                   y = c(10, 0), # add 10 cells on top and below
                                   fill = 2,)
p3 <- d6geodata::plot_qualitative_map(temp_exp_tb_sides)
p3 + p_berlin
```

### Expand On One Side With A Costum Extent

In this case we use a workaround by using a costum extent. We are adding on the left (western) side cells by using the xmin of the larger extend. Xmax, ymin, ymax are from the given extent.

```{r}
temp_exp_1_sides <- terra::extend(x = temp,
                                  y = ext(ext(temp_exp_all)[1], # xmin
                                          ext(temp)[2], # xmax
                                          ext(temp)[3], # ymin
                                          ext(temp)[4]),# ymax
                                  fill = 2)

p4 <- d6geodata::plot_qualitative_map(temp_exp_1_sides)
p4 + p_berlin
```

### Combined Plot

Here you see a combination of all plots made to compare them directly.

```{r, preview = TRUE}
(p1 + p_berlin + theme(legend.position="none") + p2 + p_berlin) / (p3 + p_berlin + p4 + p_berlin) + 
  plot_layout(guides = "collect") +
   plot_annotation(tag_levels = "1", tag_prefix = "(P", tag_suffix = ")")
```

## Rasterize Vector Data {#rasterize}

With `vect()` you can make a spatVector (a kind of shapfile) out of an extent object.
Afterwards, we can use this spatVector to rasterize the data. This works similar with an sf object.

```{r plot}
temp_vec <- terra::vect(ext(temp), crs(temp)) # spatVector out of smalest raster extent

temp_rast <- rasterize(temp_vec, # vector data
                       temp_exp_all %>% disagg(10), 
                       # raster data with larger extent, disaggregated to 100m 
                       background = 2) # set to 2 to visualize the difference

temp_rast
```

### Rasterized Plot

```{r}
d6geodata::plot_qualitative_map(temp_rast)
```

## Extract With Vector Data {#extract}

Imagine you have a track of your animal and you want to see, which type of information lays in this track. In this case, you can use your line and extract the values from a raster below. There are several ways to this but the easiest way is to use the `extract()` function form the `{terra}` package to get the values laying under the vector data.

For this, you can use any vector data (points, lines, polygons) with the same projection and extent to extract values from a raster. The `extract()` function uses the raster as first input and the vector data as second. Depending on the type of data you are using you can define a function for the extraction like min, max or mean. 'Cells' gives you the call ID, 'xy' the coordinates and 'ID' the row number of your extraction.

### Line For Extraction

For this we just creating a dummy line with coordinates laying within the raster

```{r}
pts <- tibble(x = c(4528790, 4576890), # a point out of two coordinates from the raster
              y = c(3271740, 3272140))

dummy_line <- st_as_sf(pts,             # a line made out of the point data
                       coords = c("x", "y"),
                       crs = st_crs(temp)$wkt, # using the same projection as for the template raster
                       sf_column_name = "geometry",
                       remove = FALSE) %>% 
  summarise(do_union = FALSE) %>% # this part is for creating the lines out of the points
  st_cast("LINESTRING") %>% # this as well
  mutate(id = 1:n()) # add an id
```

### Plot Of Raster And Line

```{r}
p5 <- d6geodata::plot_qualitative_map(temp_rast) + 
  geom_sf(data = dummy_line)

p5

```

### Extraction From Line

Here we finally extract the data and view the result with a table.

```{r}
ext_tab <- terra::extract(x = temp_rast, # template raster as first input
               y = dummy_line, # dummy line as second input
               fun=NULL, # here you can set a function to summarise the data directly
               method="simple", 
               cells=FALSE, # here you can get the cell number as well
               xy=FALSE, # here the xy coordinates 
               ID=TRUE # and the id in front of each row
               )
```

As you already can see on the plot, the largest part of the line lays within the purple area (1). This function can be used in cases like you need to know in which landuse class your track lays.

```{r}
table(ext_tab$layer)
```

## Visualize data with tidyterra {#tidyterra}

There are several ways to plot raster data with ggplot and one relatively new way is to do it with the package `{tidyterra}`. It has some well developed `{ggplot}` functions, but can sometimes not be combined with the base functions of `{ggplot}`. If you use this pacakge you may have to stay within the package functions.


```{r}
library(tidyterra)

ggplot() + 
  geom_spatraster(data = temp_rast) +
  p_berlin


ggplot() + 
  geom_spatraster(data = temp_rast) +
  scale_fill_whitebox_c(palette = "purple") +
  p_berlin

```


## Summary {#summary}

You learned how to extend a raster and how to extract data from a raster by using `{sf}` and `{terra}`. Of course there are several ways to do this, but this the most convenient way!!!
