---
title: "Turning equations into functions in NetLogo"
description: |
  While some programming languages, such as R, offer many native functions, others (e.g., NetLogo) offer fewer built-in options. However, users can create their own functions easily. Here, we will show the example of a Weibull density distribution function & associated cumulative density distribution function—both not yet implemented in NetLogo or NetLogo extensions.
categories:
  - NetLogo
  - rstats
  - distributions
author:
  - name: Marius Grabow
    affiliation: IZW Berlin
    affiliation_url: https://ecodynizw.github.io/EcoDynIZW
date: 11/11/2022
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


## Density distribution function

**Let's first take a look at the Weibull density distribution function:**

```{=tex}
\begin{equation}

f(x) = \frac{\gamma} {\alpha} (\frac{x-\mu}
{\alpha})^{(\gamma - 1)}\exp{(-((x-\mu)/\alpha)^{\gamma})}
\hspace{.3in}  x \ge \mu; \gamma, \alpha > 0

\end{equation}
```

### R

In R, this is already implemented:

```{r}
scale <- 3
shape <- 1

dweibull(scale,shape=shape)
```

### NetLogo

In Netlogo we can simply translate the mathematical equation into a function:

``` {.netlogo}
to-report weibull [a_scale a_shape x]

  let Wei (a_shape / a_scale ) * ((x / a_scale)^(a_shape - 1)) * exp( - ((x / a_scale)^ a_shape))

  report Wei

end
```

## Cumulative density function

**Let's also take a look at the Weibull cumulative density function:**

```{=tex}
\begin{equation}

F(x) = 1 - e^{-(x^{\gamma})} \hspace{.3in}  x \ge 0; \gamma > 0

\end{equation}
```

### R

Again, fully implemented in R already:

```{r}
scale <- 3
shape <- 1
x <- 5

pweibull(q=x, scale = scale, shape = shape)
```

### NetLogo

In Netlogo we can simply translate the mathematical equation into a function:

``` {.netlogo}
to-report weibull_cumulative [a_scale a_shape x]

  let Wei_cumu 1 - exp( - ((x / a_scale)^ a_shape))

  report Wei_cumu

end
```
